Liveness-Based Garbage Collection for Lazy Languages

### 2015-07-10 paper/paper.tex
Added a new table comparing liveness vs reachability for closures

\subsection{Copying closure using liveness vs copying closures using reachability}
To compare the  effect of using reachability to  copy closures instead
of liveness we take a couple  of programs from our benchmark suite and
execute them with both. As shown in \cred{figure-add-figure-here}, the
number of  garbage collections  in both  cases are  same but  the time
taken  varies largely.  Both these  observations can  be explained  as
follows, since we approximate the  liveness of closure to the liveness
at its  creation point, the  liveness we  get is usually  grossly over
approximated.  Using this liveness may not  give too much of a benefit
compared to  reachability as  nearly everything would  be live  at its
point of  creation.  As mentioned  in Section~\ref{sec:live-clo}, using  liveness for
closures  may  entail  multiple  traversals over  the  same  structure
consuming more time.  Therefore we believe that using reachability for
closures is  a sweet spot in  terms of performance for  liveness based
garbage collectors.
%%
%% \comment{
%% The graphs  in Figure~\ref{fig:memory-usage} show the  number of cells
%% in the heap over time for RGC and LGC---here time is measured in terms
%% of the number of cons cells allocated. In addition, they also show the
%% number of  reachable cells and the  number of cells  that are actually
%% live\footnote{This   is  statically   approximated  by   our  liveness
%%   analysis}.  Since  the programs have  different memory requirements,
%% we have tuned the size of heap for each program to ensure a reasonable
%% number of  collections. An invocation  of RGC decreases the  number of
%% cells  in heap  till  it touches  the  curve of  reachable cells.   An
%% invocation of LGC decreases the number  of heap cells to no lower than
%% the curve of live cells.

@@

In the space  of functional languages, improving  space efficiency has
been  mainly  studied  as  a compile  time  activity.  Either  through
rewriting                methods                such                as
deforestation~\cite{wadler88deforest,gill93ashort,chitil99deforest},
sharing  analysis   based  reallocation~\cite{jones89compile},  region
based  analysis~\cite{tofte98region} or  through insertion  of compile
time              nullifying              statements              such
as~\cite{inoue88analysis,lee05static,Hamilton}.    The  compile   time
marking approaches all rely on an efficient and precise alias analysis
and in the absence of it cannot provide notable improvement.

@@

%% One issue we  highlighted was that while fewer  nodes were marked (and
%% more garbage collected), sometimes  cons cells could be visited and
%% traversed  multiple times  with different  sets of  liveness  paths to
%% explore.  Further work includes  improvements to the classical copying
%% collector to reduce the cost of this.
{\color{blue}We extended the liveness based garbage collection to lazy
  languages and shown its benefit  for garbage collection in practical
  programs. We  defined context  senstive liveness analysis  that uses
  context independent summaries of functions obtained using a symbolic
  demand. We show  that obtaining precise solution  to these equations
  is undecidable in  our formulation, and hence  we safely approximate
  the result using DFAs. These DFAs are consulted by garbage collector
  to improve to collection.

  Liveness   of   closure   presented  further   challenges   to   our
  implementation.   We compared  different approximate  strategies for
  handling   liveness   of   closures,   and  found   that   a   mixed
  strategy---reachability  based  collection   of  closure  arguments,
  liveness  based  collection  of   everything  else---works  best  in
  practice as it  avoid runtime and space  overheads.  With mixed-mode
  garbage collection scheme,  we were able to collect  more cells than
  reachability based collectors at a reasonable overhead.

>>>>

\subsection{Copying closure using liveness vs copying closures using reachability}

\begin{table*}[ht!]
\caption{Comparing LGC with liveness for closures and reachability for closures}
\label{tab:compare-live-clo-reach-clo}
\centering
\begin{tabular}
{| l | r | r | r | r |}
\hline
Prog Name & Heap size & Num GC & Reachablity for closure & Liveness for closure \\ \hline
\verb@nqueens(8)@ & 10000 & 183 & 0.0087 & 0.2267\\ \hline
\verb@lambda@&10000&11&0.0010&4.9922\\ \hline
\verb@lcss@&10000&1&0.00008&34.6684\\ \hline
\end{tabular}
\end{table*}
To compare the  effect of using reachability to  copy closures instead
of liveness we take a couple  of programs from our benchmark suite and
execute       them      with       both.        As      shown       in
Table~\ref{tab:compare-live-clo-reach-clo},  the   number  of  garbage
collections in both cases are same but copying closures using liveness
takes a lot more time compared to reachability based copying.

The number  of garbage  collections is same  since we  approximate the
liveness of  a closure  to the  liveness at  its creation  point.  The
liveness  we get  is usually  grossly over  approximated.  Using  this
liveness may not  give too much of a benefit  compared to reachability
as nearly everything would be live at its point of creation.

As  mentioned   in  Section~\ref{sec:live-clo},  using   liveness  for
closures  may  entail  multiple  traversals over  the  same  structure
consuming more  time.  The  advantage of avoiding  multiple traversals
can be clearly seen from the lcss example where a single GC takes less
than a millisecond for reachability  based closure collection where as
the  liveness  based collection  is  of  the  order of  seconds.  Lazy
languages tend  to have a  lot of sharing  and hence using  a liveness
based collection for closures is impractical.

Therefore we believe  that using reachability for closures  is a sweet
spot   in   terms   of   performance  for   liveness   based   garbage
collectors.  For the  remainder  of  the section  all  LGC results  we
discuss uses  a mixed-mode collector--liveness for  evaluated data and
reachability for closures.
%%
%% \comment{
%% The graphs  in Figure~\ref{fig:memory-usage} show the  number of cells
%% in the heap over time for RGC and LGC---here time is measured in terms
%% of the number of cons cells allocated. In addition, they also show the
%% number of  reachable cells and the  number of cells  that are actually
%% live\footnote{This   is  statically   approximated  by   our  liveness
%%   analysis}.  Since  the programs have  different memory requirements,
%% we have tuned the size of heap for each program to ensure a reasonable
%% number of  collections. An invocation  of RGC decreases the  number of
%% cells  in heap  till  it touches  the  curve of  reachable cells.   An
%% invocation of LGC decreases the number  of heap cells to no lower than
%% the curve of live cells.

@@

In the space  of functional languages, improving  space efficiency has
been  mainly  studied  as  a compile  time  activity.  Either  through
rewriting                methods                such                as
deforestation~\cite{wadler88deforest,gill93ashort,chitil99deforest},
sharing  analysis   based  reallocation~\cite{jones89compile},  region
based  analysis~\cite{tofte98region}, or  through insertion  of compile
time              nullifying              statements              such
as~\cite{inoue88analysis,lee05static,Hamilton}.    The  compile   time
marking approaches all rely on an efficient and precise alias analysis
and in the absence of it cannot provide notable improvement.

@@

%% One issue we  highlighted was that while fewer  nodes were marked (and
%% more garbage collected), sometimes  cons cells could be visited and
%% traversed  multiple times  with different  sets of  liveness  paths to
%% explore.  Further work includes  improvements to the classical copying
%% collector to reduce the cost of this.
{\color{blue}We extended the liveness based garbage collection to lazy
  languages and shown its benefit  for garbage collection in practical
  programs. We  defined context  senstive liveness analysis  that uses
  context independent summaries of functions obtained using a symbolic
  demand. We show  that obtaining precise solution  to these equations
  is undecidable in  our formulation, and hence  we safely approximate
  the result using DFA. These DFA are consulted by the garbage collector
  to improve  collection.

  Liveness   of   closure   presented  further   challenges   to   our
  implementation.   We compared  different approximate  strategies for
  handling   liveness   of   closures,   and  found   that   a   mixed
  strategy---reachability  based  collection   of  closure  arguments,
  liveness  based  collection  of   everything  else---works  best  in
  practice as it  avoids runtime and space  overheads.  With mixed-mode
  garbage collection scheme,  we were able to collect  more cells than
  reachability based collectors at a reasonable overhead.
